High-Dimensional Regularized Discriminant Analysis

### 2015-10-21 reports/paper-hdrda.Rnw
New description of sim design: rough for now.

\author{John A. Ramey, Caleb K. Stein, and Dean M. Young}

@@

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier and then compare its computational runtime with an
implementation of the \emph{RDA} classifier from the {\tt klaR} R package to
illustrate that the \emph{HDRDA} classifier greatly reduces computation when $p
\gg N$. We then study the classification performance of the \emph{HDRDA}
classifier on three artificial and four real high-dimensional data sets. We
demonstrate that the \emph{HDRDA} classifier often attains superior
classification accuracy to several recent classifiers designed for small-sample,
high-dimensional data from \cite{Tong:2012hw}, \cite{Witten:2011kc},
\cite{Pang:2009ik}, \cite{Guo:2007te}, and \cite{Dudoit:2002ev}. We also include
as benchmarks support vector machines (\emph{SVM}) and $k$-nearest neigbors
($k$-NN) in our classification study.\footnote{TODO: If we change any of the
  classifiers in the simulations, they should be updated here.}

@@

In this section, we describe four high-dimensional microarray data sets and
compare our proposed classifier with five classifiers recently proposed for
small-sample, high-dimensional data. Our competitors include diagonal linear
discriminant analysis from \cite{Dudoit:2002ev}, shrunken centroids regularized
discriminant analysis from \cite{Guo:2007te}, and penalized linear discriminant
analysis from \cite{Witten:2011kc}. We also consider two modifications of
diagonal linear discriminant analysis from \cite{Tong:2012hw} and
\cite{Pang:2009ik}, where the former employs an improved mean estimator and the
latter utilizes an improved variance estimator. Below, we refer to each
classifier by the first author's surname.\footnote{TODO: Consider adding a
  heatmap of error rates of a single $(\lambda, \gamma)$ grid.}

We evaluated the classification performance of each classifier by randomly
partitioning the data set under consideration such that $2/3$ of the
observations were allocated as training data and the remaining $1/3$ of the
observations were allocated as a test data set. For each random partition, we
applied the feature-selection method from \cite{Dudoit:2002ev} to the training
data to choose the $d$ features having the largest ratio of their between-group
to within-group sums of squares.  We then filtered the test data set with the
corresponding $d$ features. After training each classifier from the filtered
training data, we classified the filtered test data sets and computed the
proportion of mislabeled test observations to estimate the classification error
rate for each classifier. Repeating this process 1000 times, we computed the
average of the error-rate estimates for each classifier.

We used version 3.1.0 of the open-source statistical software {\tt R} for the
classification study. We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively, on
CRAN. Because the {\tt rda} package does not perform the authors' ``Min-Min''
rule automatically, we applied this rule in our {\tt R} code. We also utilized
our implementations of the Dudoit, Pang, and Tong classifiers available in the
{\tt sparsediscrim} package. We used the default settings for each classifier
with the exception that we explicitly set prior probabilities as equal if the
classifier allowed the option. For the \emph{HDRDA} classifier in
\eqref{eq:hdrda-decomposed}, we examine the classification performance of two
models. For the first \emph{HDRDA} model, we set $\alpha_k = 1$, $k = 1, \ldots,
K$, so that the covariance-matrix estimator \eqref{eq:hdrda-cov} employed
resembled \eqref{eq:ridge-estimator}. We estimated $\lambda$ from a grid of 21
equidistant candidate values between 0 and 1, inclusively. Similarly, we
estimated $\gamma$ from a grid consisting of the values $10^{-1}, \ldots, 10^4$,
and $10^5$. We selected optimal estimates of $\lambda$ and $\gamma$ using
$V$-fold cross-validation with $V = 10$ \citep{Hastie:2008dt}. For the second
model, we set $\alpha_k = 1 - \gamma$, $k = 1, \ldots, K$, to resemble
\citeauthor{Friedman:1989tm}'s parameterization. We estimated both $\lambda$ and
$\gamma$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Here, we provide a concise description of each high-dimensional
data set examined in our classification study.

\subsection{\cite{Burczynski:2006ik} Data Set}

@@

\subsection{\cite{Nakayama:2007fl} Data Set}

@@

\subsection{\cite{Shipp:2002ka} Data Set}

@@

\subsection{\cite{Singh:2002fh} Data Set}

>>>>

\author{John A. Ramey, Caleb K. Stein, Phil D. Young, and Dean M. Young}

@@

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier and then compare its computational runtime with an
implementation of the \emph{RDA} classifier from the {\tt klaR} R package to
illustrate that the \emph{HDRDA} classifier greatly reduces computation when $p
\gg N$. We then study the classification performance of the \emph{HDRDA}
classifier on three artificial and four real high-dimensional data sets. We
demonstrate that the \emph{HDRDA} classifier often attains superior
classification accuracy to several recent classifiers designed for small-sample,
high-dimensional data from \cite{Tong:2012hw}, \cite{Witten:2011kc},
\cite{Pang:2009ik}, and \cite{Guo:2007te}. We also include as a benchmark the
random forest from \cite{Breiman:2001fb} because \cite{FernandezDelgado:2014ul}
has concluded that the random forest is often superior to other classifiers in
benchmark studies.

@@

In this section, we compare our proposed classifier with four classifiers
recently proposed for small-sample, high-dimensional data along with the
random-forest classifier from \cite{Breiman:2001fb} as a benchmark. Our
competitors include shrunken centroids regularized discriminant analysis from
\cite{Guo:2007te} and penalized linear discriminant analysis from
\cite{Witten:2011kc}. We also consider two modifications of diagonal linear
discriminant analysis from \cite{Tong:2012hw} and \cite{Pang:2009ik}, where the
former employs an improved mean estimator and the latter utilizes an improved
variance estimator. Below, we refer to each classifier by the first author's
surname.\footnote{TODO: Consider adding a heatmap of error rates of a single
  $(\lambda, \gamma)$ grid.}

We used version 3.2.2 of the open-source statistical software {\tt R} for the
classification study. We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively.
Because the {\tt rda} package does not perform the authors' ``Min-Min'' rule
automatically, we applied this rule within our {\tt R} code. We also utilized
the implementations of the Pang and Tong classifiers available in the {\tt
  sparsediscrim} package. We used the implementation of the random-forest
classifier from the {\tt randomForest} package. For each classifer we explicitly
set prior probabilities as equal if applicable. All other classifier options
were set to their default settings.

For the \emph{HDRDA} classifier in \eqref{eq:hdrda-decomposed}, we examine the
classification performance of two models. For the first \emph{HDRDA} model, we
set $\alpha_k = 1$, $k = 1, \ldots, K$, so that the covariance-matrix estimator
\eqref{eq:hdrda-cov} employed resembled \eqref{eq:ridge-estimator}. We estimated
$\lambda$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Similarly, we estimated $\gamma$ from a grid consisting of the
values $10^{-1}, \ldots, 10^4$, and $10^5$. We selected optimal estimates of
$\lambda$ and $\gamma$ using $V$-fold cross-validation with $V = 10$
\citep{Hastie:2008dt}. For the second model, we set $\alpha_k = 1 - \gamma$, $k
= 1, \ldots, K$, to resemble \citeauthor{Friedman:1989tm}'s parameterization. We
estimated both $\lambda$ and $\gamma$ from a grid of 21 equidistant candidate
values between 0 and 1, inclusively.

\subsection{Simulation Study}

We compared the \emph{HDRDA} classifer to the five competing classifiers in a
simulaton study. We simulated $K = 3$ $p$-dimensional multivariate normal
populations with a common positive-definite covariance matrix having a
block-diagonal covariance structure, which is intended to resemble
gene-expression data. That is, within each block of pathways, genes are
positively correlated and that this correlation decays as a function of the
distance between any two genes. Thus, we simulated $K = 3$ multivariate normal
populations with the common positive-definite covariance matrix

\begin{align*}
  \bm\Sigma = \bm\Sigma^{(\rho)} \oplus \bm\Sigma^{(-\rho)} \oplus \ldots \oplus \bm\Sigma^{(\rho)},
\end{align*}

where the $(i,j)$th entry of $\bm\Sigma^{(\rho)} \in \mathbb{R}_{25 \times 25}$ is

\begin{align*}
\bm\Sigma_{ij}^{(\rho)} = \{ \rho^{|i - j|} \}_{1 \le i,j \le 25},
\end{align*}

altering positive and negative values of $\rho$ are used from block to block. In
our simulation study, we considered three different values of $\rho$, 0.1, 0.5,
and 0.9, to explore the effect of autocorrelation on classification accuracy,
especially given that the classifiers from \cite{Tong:2012hw} and
\cite{Pang:2009ik} assume independence among features.

Furthermore, we set the population of mean of class 1 to be a $p$-dimensional
vector of zeros. Next, we set class 2's population mean to a vector equaling 1/2
for the first 50 features and 0 after ($\mu_j = 1/2$ if $j \le 50$ and $\mu_j =
0$ if $j > 50$). Similarly, we set class 3's population mean to a vector equaling -1/2
for the first 50 features and 0 after ($\mu_j = -1/2$ if $j \le 50$ and $\mu_j =
0$ if $j > 50$). The three populations differ in their mean vectors in the first
50 features corresponding to the first two blocks. No difference in the means
occurs in the remaining blocks.

The block-diagonal simulation design considered is widely used within the
high-dimensional classification literature, including the studies by
\cite{Ramey:2013ji}, \cite{Witten:2011kc}, and \cite{Guo:2007te}. While the
design is indeed standard, the simulation configuration is based on an ideal
scenario where no artifacts, often observed in real data, are present. As a
result, we wished to investigate the effect of outliers on the high-dimensional
classifiers. To do this, we generalized the block-diagonal simulation
configuration by sampling from a $p$-dimensional multivariate
$\epsilon$-contaminated normal distribution, where $\epsilon in [0, 1]$. For
$\epsilon = 0$, we have the benchmark block-diagonal simulations considered in
the literature. As $\epsilon$ is increased, the number of outliers is increased
(on average). We consider the values of $\epsilon = 0, 0.05, 0.1, 0.15,
0.2$. For the contamination simulation, we define the probability density
function of the $k$th class to be

\begin{align*}
  p(\bm x | \omega_k) = (1 - \epsilon) N_p(\bm x | \bm \mu_k, \bm \Sigma) + \epsilon N_p(\bm x | \bm \mu_k, 100 * \bm \Sigma)
\end{align*}

From each of the $K=3$ populations, we sampled 25 training observations ($n_k =
25$ for all $k$) and 1000 test observations. After training each classifier on
the training data, we classified the test data sets and computed the proportion
of mislabeled test observations to estimate the classification error rate for
each classifier. Repeating this process 250 times, we computed the average of
the error-rate estimates for each classifier. We allowed the number of features
to vary from $p = 100$ to $p = 500$ in increments of 100 to examine the increase
in dimension while maintaining a small sample size.

Despite the claims of \cite{FernandezDelgado:2014ul} that the random forest
classifier is typically superior, we observed that the classifiers designed for
high-dimensional data were indeed superior.

\subsection{Application to Gene Expression Data}

We compared the \emph{HDRDA} classifer to the five competing classifiers on four
benchmark gene-expression microarray data sets. We evaluated the classification
performance of each classifier by randomly partitioning the data set under
consideration such that $2/3$ of the observations were allocated as training
data and the remaining $1/3$ of the observations were allocated as a test data
set. After training each classifier on the training data, we classified the test
data sets and computed the proportion of mislabeled test observations to
estimate the classification error rate for each classifier. Repeating this
process 250 times, we computed the average of the error-rate estimates for each
classifier. Here, we provide a concise description of each high-dimensional
data set examined in our classification study.

\subsubsection{\cite{Burczynski:2006ik} Data Set}

@@

\subsubsection{\cite{Nakayama:2007fl} Data Set}

@@

\subsubsection{\cite{Shipp:2002ka} Data Set}

@@

\subsubsection{\cite{Singh:2002fh} Data Set}
